Context-dependent consequences of lagged effects in demographic models

### 2025-10-28 docs/ipm_comparison_ms.qmd
edits figs and bib

Current environmental conditions can have large and immediate effects on the growth, survival, or reproduction of long-lived organisms. There is also mounting evidence, however, for intervals that range from months to years between environmental conditions and the resulting changes in demographic vital rates. These *Lagged Effects*, also known as *Delayed Life-history Events* (i.e., DLHEs) [@beckermanPopulationDynamicConsequences2002], can simultaneously affect an entire cohort (e.g., juveniles hatching during a period of scarcity will all have delayed maturation and lower lifetime fecundity) or only a subset of the population (e.g., cold temperatures in one year lead to reduced flowering by potentially reproductive individuals in the next). In addition, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @crileyYearProductionHigh1994]. A drought during the early stages of gestation or floral bud formation, for example, might have a much larger impact on the number of fruits or offspring produced than one at a later stage of the reproductive process. The delay or intensity of lagged effects can also be location- or habitat-specific, with individuals in some sites or habitats buffered against, or able to recover more quickly from, the delayed effects of environmental variation.

Because Lagged Effects are often directly linked to reproduction and survival, it has been argued they could have major but underappreciated consequences for population dynamics [@beckermanPopulationDynamicConsequences2002]. While recent studies suggest this could indeed the case [e.g., @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018; @williamsLifeHistoryEvolution2015], broader efforts to test this hypothesis have face a number of obstacles [@metcalfStatisticalModellingAnnual2015]. Empirical tests of putative lagged effects are rare because these are challenging to design, implement, and maintain [@kussEvolutionaryDemographyLonglived2008a]. While one could conceivably use observational data to detect lagged effects, doing so requires long-term data on both the putative lagged effect (e.g., reduced probability of flowering) and its potential environmental drivers (e.g., drought during early floral development), and such coupled data sets are rare [sensu @eversLaggedDormantSeason2021]. In addition, the methods for both identifying lagged effects and modeling their demographic impacts can be challenging to implement. For example, many of the potentially applicable statistical methods have stringent assumptions and data requirements rarely met by ecological data [@metcalfStatisticalModellingAnnual2015], while the including complex biological processes in demographic models can render them less tractable or amenable to analysis using currently available tools. Addressing these challenges is a major undertaking; the value of doing so will depend on the effort required vs. the potential consequences of failing to consider lagged effects --- consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.

Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellnerIntegralProjectionModels2006; @reesBuildingIntegralProjection2014; @reesIntegralProjectionModels2009]. Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011]. Mathematical and statistical advances [e.g., @williams2012avoiding; @brooksStatisticalModelingPatterns2019] have rapidly expanded the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016]. Here we investigate how including lagged effects in Integral Projection Models influences projections of $\lambda$ and population structure.

We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat. We parameterized three classes of Integral Projection Models --- a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates --- for populations in two habitat types (i.e., continuous forest vs. forest fragments). Based on previous studies [@brunaArePlantPopulations2003; @brunaDemographicEffectsHabitat2005; @brunaHabitatFragmentationDemographic2002] and demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990] we predicted that: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) $\lambda$ is lower for Forest Fragment than Continuous Forest populations regardless of model, but the difference between the two habitats is greatest for the IPM with lagged effects.

@@

The rainy season in our sites is typically from late December through late May. *Heliconia acuminata* in our site begin flowering early in the rainy season and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999]. On average, plots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).

@@

The second is sub-kernel $R(z^\prime)$, which describes the survival of seedlings established in year $t$ and their size when entering the mature plant population in year $t+1$ (@eq-R):

@@

Note that in @eq-R both the probability that new seedlings survive their first year, $s_s$, and their size at the end of this year, $G(z^\prime, z)$, are size-independent. IPMs can include transitions between individuals from a discrete state to a continuous one [i.e., from 'seedling' to 'mature plant of size $z'$', @ellnerDatadrivenModellingStructured2016]; we treat seedlings as a distinct and discrete category because they have lower survival and growth in their first year than comparably sized plants [@brunaArePlantPopulations2003; @scottDelayedEffectsClimate2022].\

@@

Both the probability that a mature plant will flower, $p_f(z)$, and the number of seeds a flowering plant will produce, $f(z)$, are size-dependent. All seeds germinate and establish as seedlings with probability $g$.\
<!-- `r version$version.string`  -->

@@

**(2) Stochastic IPMs:** To include temporal stochasticity in our IPMs we included a random effect of year. This was done using a factor--smooth interaction that allowed the functional form of the relationship between plant size and vital rates to vary among transition years (@fig-lifecycle). We generated kernels for every transition year using the long-term survey data [@brunaDemographyUnderstoryHerb2023] and random smooths for year. We then randomly selected one of these sets of kernels to use in each iteration of the IPM. This procedure is equivalent to *'kernel resampling'* [*sensu* @metcalfStatisticalModellingAnnual2015] or matrix selection for matrix population models [@caswell2001].\

@@

##### Comparison of $\lambda$ in Continuous Forest and Forest Fragments.

@@

##### Population Structure.

@@

Including lagged effects in IPMs also resulted in significantly less predictable projections of population structure, both within (@fig-pop-states a) and across habitats (@fig-pop-states b, @tbl-vartest). This variability is particularly notable in forest fragments (@fig-marginal, @tbl-meanvars), where we have previously shown lagged effects have very large impacts on growth and survival [@scottDelayedEffectsClimate2022]. The deterministic IPM for Continuous Forest projected slightly more of the smallest and largest plants than the one for Forest Fragments. In other words, populations in forest fragments had proportionately less recruitment and fewer individuals growing into the larger, reproductive size classes [@brunaDemographicEffectsHabitat2005]. This shift towards intermediately sized, pre-reproductive plants in forest fragments is even more dramatic when using the parameter-resampled stochastic IPM (@fig-pop-states b).\

What are the implications of these results for the use of IPMs to study plant demography? First, they suggest that failing to include lagged effects in models could significantly underestimate the growth rate of populations. While relative differences in projections of $\lambda$ may be sufficient for some tests of theory, the actual value of projections may be critical in conservation or management settings. For example, they could be used to identify which populations face the greatest risk of extinction (e.g., $\lambda=0.99$ vs. $\lambda=1.01$) or --- because even small differences in $\lambda$ can have major short-term consequences for population dynamics (e.g., $\lambda=1.01$ vs. $\lambda=1.02$) - be used to chose among management strategies. Moreover, including lagged effects in models could be especially critical for studies assessing the effects of climate change, given many DLHE are climate driven. That said, there is a cost to including lagged effects in demographic models, even when doing so is conceptually appropriate or if the outcome is potentially more accurate. While Deterministic and Kernel-resampled Stochastic IPMs took only \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Stochastic IPMs with using parameter-resampled kernels and lagged effects took \~`r spar_time` min. This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower for Generalized Additive Models with 2D smooths because of the much higher number of knots than for the GAMs used in Determinsitic and Kernel-resampled IPMs. Advances in computational power and access to high-performance computing resources could lower this cost.

That said, computational power cannot compensate for limited data. Detecting lagged effects and evaluating their consequences requires long-term demographic data --- data that are only available for a relatively small number of species, few of which are in the tropics. The increasing evidence that lagged effects are ubiquitous, and that they can have major demographic impacts, underscores the need to support the collection of such long-term data, the complementary development of experimental and statistical approaches to disentangling lagged effects, and community driven efforts to identify priority or model systems for in-depth investigation.\

@@

```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "**(A)** The change over time in the the proportion of _Heliconia acuminata_ populations in different size/stage classes when simulating population dyammics in Continuous Forest (CF) or Forest Fragment (FF) with three different Integral Projection Models. Results are shown for the first 250 iterations of populations; for the criteria used to define the size categories see @tbl-plantcats. **(B)** The relative proportion of the population in each size class (FF vs. CF) for 250 iterations of each IPM model. Note that this excludes transient dynamics (iterations 1-30). Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class."
# note flowering prob for rp1 <0.25 but for rp2>0.2???'
# include _seedlings_ (a discrete category in the IPMs), _pre-reproductive 1_ (log(size) = 0–2.5, average survival < 0.9, probability of flowering \\approx 0), _pre-reproductive 2_ (log(size) = 2.5–4.5, average survival probabilty > 0.8, probability of flowering \\approx 0), _reproductive 1_ (log(size) = 4.5–6, average survival probability >0.95, flowering probability < 0.25), and _reproductive 2_ (log(size) = 6+, average survival probability >0.95, flowering probability > 0.2)
knitr::include_graphics(here(fig_pop_states), rel_path = TRUE)
```

@@

```{r #fig-lambdas,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "Distribution of 900 values of $\\lambda$ projected with (A) Stochastic, kernel-resampled IPMs and (B) Stochastic, paramenter-resampled IPMs. IPMs were used to project $\\lambda$ for both Continuous Forest (above, in green) and Forest Fragments (below, in gray). The solid line indicates the mean value of $\\lambda$, the dashed line indicates the value of $\\lambda$ in that habitat projected with Deterministic IPMs."

@@

```{r #fig-marginal,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "The relative proportion of the population in each size class (FF vs. CF) for each of 250 iterations of the kernel-resampled (dark shading) and parameter-resampled IPM models (light shading). Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class; the marginal plots indicate the distribution of these relative proportions for each class of IPM in each habitat. Note that both the scatterplots and marginal plots exclude transient dynamics (iterations 1-30)."

@@

```{r aic, results='asis'}
#| label: tbl-aic
#| tbl-cap: "Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
#| tbl-pos: H
#| 
aic_tbl_df %>%
  group_by(habitat, model) %>% 
  # mutate(rank = row_number()) %>% 
  ungroup() %>% 
  mutate(habitat = toupper(habitat),
         env_effect_incl = ifelse(
           env_effect_incl != "DLNM",
           to_sentence_case(as.character(env_effect_incl)),
           as.character(env_effect_incl)
         )
  ) %>% 
  mutate(
    model = fct_relevel(
      model,
      c("surv", "size", "flwr", "surv_sdlg", "size_sdlg")
    ),
    model = fct_recode(
      model,
      "Survival"          = "surv",
      "Growth"            = "size",
      "Flowering"         = "flwr",
      "Seedling survival" = "surv_sdlg",
      "Seedling growth"   = "size_sdlg"
    )
  ) %>%

@@

  arrange(habitat, model) %>% 
  mutate(habitat=c(rep("Continuous Forest",15),rep("Forest Fragments",15))) %>% 
  mutate(model=(rep(c("Survival","","","Growth","","","Flowering","","","Seedling survival","","","Seedling growth","",""),2)))  %>%
  select(
    "Habitat" = "habitat",
    "Vital Rate" = "model",
    "Effect of Environment" = "env_effect_incl",
    #"Model" = "RHS",
    "edf" = "df",
    "\u0394AIC" = "dAIC"
  )  %>% 
  mutate(Habitat = str_replace_all(
    Habitat,
    c(
      "Continuous Forest" = "",
      "Forest Fragments" = ""
      )
  )) %>% 
  kable(digits = 2,
  format = "latex",
  align = "cllcc",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  linesep = ""
) %>%
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 10,
    position = "center"
  ) %>% 
# %>% 
  # collapse_rows(columns = 1:2, valign = "top")
  column_spec(1, width = "10em") %>% 
  pack_rows("Continuous Forest", 1, 15) %>%
  pack_rows("Forest Fragments", 16, 30)

@@

lambda_table %>% 
  mutate(habitat = toupper(habitat)) %>% 
  mutate(across(c(est, lower, upper), ~format(round(.,4)))) %>% 
  # mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
  mutate(CI = glue::glue("({lower}, {upper})")) %>%
  mutate(ipm = str_replace_all(
    ipm,
    c(
      "det" = "Deterministic",
      "stoch" = "Stochastic, kernel-resampled",
      "dlnm" = "Stochastic, parameter-resampled"
    )
  )) %>% 
  # separate(lambda,c("lambda","CI"),sep=" [(]") %>% 
  select(IPM = ipm, 
         Habitat = habitat, 
         # "$\\lambda$" = lambda) %>% 
         "$\\lambda$" = est,
         "(Lower, Upper 95\\% CI)"=CI) %>%
  mutate(IPM = str_replace_all(
    IPM,
    c(
      "Deterministic" = "",
      "Stochastic, kernel-resampled" = "",
      "Stochastic, parameter-resampled" = ""
      )
  )) %>%  
  kable(digits = 2,
  format = "latex",
  align = "lccc",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  linesep = ""
) %>% 
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  ) %>% 
  # collapse_rows(columns = 1:2, valign = "top")
  column_spec(1, width = "15em") %>% 
  pack_rows("Deterministic", 1, 2) %>%
  pack_rows("Stochastic, kernel resampled", 3, 4) %>% 
  pack_rows("Stochastic, parameter-resampled", 5, 5)

@@

stats_table %>%
  # select(-`Vital Rate`) %>%
  rename(`Stage`=`Vital Rate`) %>% 
  mutate(`Stage`= " ") %>% 
  # mutate(Comparison="KR v PR") %>% 
  # relocate(Comparison,.before=1) %>% 
kable(digits = 5,
  format = "latex",
  align = "lcrrcr",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  # caption = "Test comparing variances of vital rates in stochastic ipms.",
  linesep = ""
) %>%
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  ) %>%
  # collapse_rows(columns = 1:2, valign = "top")
  column_spec(1, width = "8em") %>%
  # collapse_rows(columns = 1:2,valign="top")
  pack_rows("Seedling", 1, 2) %>%
  pack_rows("Pre-reproductive 1", 3, 4) %>%
  pack_rows("Pre-reproductive 2", 5, 6) %>%
  pack_rows("Reproductive 1", 7, 8) %>%
  pack_rows("Reproductive 2", 9, 10)

>>>>

Current environmental conditions can have immediate effects on the growth, survival, or reproduction of long-lived organisms. However, there is mounting evidence for potential delays of months or even years between current environmental conditions and resulting changes in demographic vital rates. These *Lagged Effects*, also known as *Delayed Life-history Events* [i.e. DLHEs, @beckermanPopulationDynamicConsequences2002], can simultaneously affect an entire cohort (e.g., all juveniles hatching during a period of scarcity will have delayed maturation) or only a subset of the population (e.g., cold winter temperatures reduce flowering by small trees but not large ones). Moreover, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @crileyYearProductionHigh1994]. For example, a drought during the early stages of gestation or floral bud formation might have a much larger impact on the number of fruits or offspring produced than one at a later stage of the reproductive process. The delay or intensity of lagged effects can also be location- or habitat-specific, with individuals in some sites or habitats buffered against, or able to recover more quickly from, the delayed effects of environmental variation.

Because Lagged Effects are often directly linked to reproduction and survival, it has been argued they could have major but underappreciated consequences for population dynamics [@beckermanPopulationDynamicConsequences2002]. While recent studies suggest this could indeed the case [e.g., @williamsLifeHistoryEvolution2015; @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018], efforts to test this hypothesis are rare. [@metcalfStatisticalModellingAnnual2015]. In part this is because empirical tests of putative lagged effects are are challenging to design, implement, and maintain [@kussEvolutionaryDemographyLonglived2008a]. While one could conceivably use observational data to detect lagged effects, doing so requires long-term data on both the putative lagged effect (e.g., reduced probability of flowering) and its potential environmental drivers (e.g., drought during early floral development), and such coupled data sets are rare [sensu @eversLaggedDormantSeason2021]. In addition, the methods for both identifying lagged effects and modeling their demographic impacts can be challenging to implement. For example, many of the potentially applicable statistical methods have stringent assumptions and data requirements rarely met by ecological data [@metcalfStatisticalModellingAnnual2015], while the including complex biological processes in demographic models can render them less tractable or amenable to analysis using currently available tools. Addressing these challenges is a major undertaking, and the value of doing so will depend on the effort required vs. the potential consequences of failing to consider lagged effects --- consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.

Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellnerIntegralProjectionModels2006; @reesBuildingIntegralProjection2014; @reesIntegralProjectionModels2009]. Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011]. Mathematical and statistical advances [e.g., @williams2012avoiding; @brooksStatisticalModelingPatterns2019] have rapidly expanded the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016]. Here we test predictions for how including lagged effects in Integral Projection Models will influence projections of $\lambda$ and population structure.

We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat (i.e., continuous forest vs. forest fragments). We parameterized three classes of Integral Projection Models --- a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates --- for populations in these two habitat types to test the following predictions from demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990]: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) $\lambda$ is lower for Forest Fragment than Continuous Forest populations regardless of model, but the difference between the two habitats is greatest for the IPM with lagged effects.

@@

The rainy season in our sites is typically from late December through late May. *Heliconia acuminata* in our site begin flowering early in the rainy season and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999]. On average, plots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF: median = 788, range = 201-1549; 1-ha: median = 339, range = 297-400).

@@

\noindent{The} second is sub-kernel $R(z^\prime)$, which describes the survival of seedlings established in year $t$ and their size when entering the mature plant population in year $t+1$ (@eq-R):

@@

\noindent{Note} that in @eq-R both the probability that new seedlings survive their first year, $s_s$, and their size at the end of this year, $G(z^\prime, z)$, are size-independent. IPMs can include transitions between individuals from a discrete state to a continuous one [i.e., from 'seedling' to 'mature plant of size $z'$', @ellnerDatadrivenModellingStructured2016]; we treat seedlings as a distinct and discrete category because they have lower survival and growth in their first year than comparably sized plants [@brunaArePlantPopulations2003; @scottDelayedEffectsClimate2022].\

@@

\noindent{Both} the probability that a mature plant will flower, $p_f(z)$, and the number of seeds a flowering plant will produce, $f(z)$, are size-dependent. All seeds germinate and establish as seedlings with probability $g$.\
<!-- `r version$version.string`  -->

@@

**(2) Stochastic IPMs:** To include temporal stochasticity in our IPMs we included a random effect of year. This was done using a factor--smooth interaction that allowed the functional form of the relationship between plant size and vital rates to vary among transition years (@fig-lifecycle). We generated kernels for every transition year using the long-term survey data [@brunaDemographyUnderstoryHerb2023] and random smooths for year. We then randomly selected one of these sets of kernels to use in each iteration of the IPM. This procedure is equivalent to *'kernel resampling'* [*sensu* @metcalfStatisticalModellingAnnual2015] or matrix selection for matrix population models [@caswell2001; @boyceDemographyIncreasinglyVariable2006a].\

@@

### Comparison of $\lambda$ in Continuous Forest and Forest Fragments

@@

### Population Structure

@@

Including lagged effects in IPMs also resulted in significantly less predictable projections of population structure, both within (@fig-pop-states A) and across habitats (@fig-pop-states B, @tbl-vartest). This variability is particularly notable in forest fragments (@fig-marginal, @tbl-meanvars), where we have previously shown lagged effects have very large impacts on growth and survival [@scottDelayedEffectsClimate2022]. The deterministic IPM for Continuous Forest projected slightly more of the smallest and largest plants than the one for Forest Fragments. In other words, populations in forest fragments had proportionately less recruitment and fewer individuals growing into the larger, reproductive size classes [@brunaDemographicEffectsHabitat2005]. This shift towards intermediately sized, pre-reproductive plants in forest fragments is even more dramatic when using the parameter-resampled stochastic IPM (@fig-pop-states B).\

What are the implications of these results for the use of IPMs to study plant demography? First, they suggest that failing to include lagged effects in models could significantly underestimate the growth rate of populations. While relative differences in projections of $\lambda$ may be sufficient for some tests of theory, the actual value of projections may be critical in conservation or management settings. For example, they could be used to identify which populations face the greatest risk of extinction (e.g., $\lambda=0.99$ vs. $\lambda=1.01$) or, because even small differences in $\lambda$ can have major short-term consequences for population dynamics (e.g., $\lambda=1.01$ vs. $\lambda=1.02$), be used to chose among alternative management strategies. Moreover, including lagged effects in models could be especially critical for studies assessing the effects of climate change, given many DLHE are climate driven.

That said, there is a cost to including lagged effects in demographic models, even when doing so is conceptually appropriate or if the outcome is potentially more accurate. While Deterministic and Kernel-resampled Stochastic IPMs took only \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Stochastic IPMs with using parameter-resampled kernels and lagged effects took \~`r spar_time` min. This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower for Generalized Additive Models with 2D smooths because of the much higher number of knots than for the GAMs used in Determinsitic and Kernel-resampled IPMs. Advances in computational power and access to high-performance computing resources could lower this temporal and computational cost.

It is important to emphasize, however, that computational power cannot compensate for limited data. Detecting lagged effects and evaluating their consequences requires long-term demographic data --- data that are only available for a relatively small number of species, few of which are in the tropics [@brunaHabitatFragmentationPlant2009; @romerPlantDemographicKnowledge2024]. The increasing evidence that lagged effects are ubiquitous, and that they can have major demographic impacts, underscores the need to support the collection of such long-term data, the complementary development of experimental and statistical approaches to disentangling lagged effects, and community driven efforts to identify priority or model systems for in-depth investigation.\

@@

```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "**(A)** The change over time in the the proportion of _Heliconia acuminata_ populations in different size/stage classes when simulating population dyammics in Continuous Forest (CF) or Forest Fragment (FF) with three different integral projection models (e.g., IPMs). Results are shown for the first 250 iterations of populations; for the criteria used to define the size categories see @tbl-plantcats. **(B)** The relative proportion of the population in each size class (FF vs. CF) for 250 iterations of each IPM model. Note that this excludes transient dynamics (iterations 1-30). Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class."
# note flowering prob for rp1 <0.25 but for rp2>0.2???'
# include _seedlings_ (a discrete category in the IPMs), _pre-reproductive 1_ (log(size) = 0–2.5, average survival < 0.9, probability of flowering \\approx 0), _pre-reproductive 2_ (log(size) = 2.5–4.5, average survival probabilty > 0.8, probability of flowering \\approx 0), _reproductive 1_ (log(size) = 4.5–6, average survival probability >0.95, flowering probability < 0.25), and _reproductive 2_ (log(size) = 6+, average survival probability >0.95, flowering probability > 0.2)
knitr::include_graphics(here(fig_pop_states), rel_path = TRUE)
```

@@

```{r #fig-lambdas,  echo=FALSE, out.width="100%",fig.align='center'}
#| fig.cap = "Distribution of 900 values of $\\lambda$ projected with (A) Stochastic, kernel-resampled IPMs and (B) Stochastic, paramenter-resampled IPMs. IPMs were used to project $\\lambda$ for both Continuous Forest (above, in green) and Forest Fragments (below, in gray). The solid line indicates the mean value of $\\lambda$, the dashed line indicates the value of $\\lambda$ in that habitat projected with Deterministic IPMs."

@@

```{r #fig-marginal,  echo=FALSE, out.width="75%",fig.align='center'}
#| fig.cap = "The relative proportion of the population in each size class (FF vs. CF) for each of 250 iterations of the kernel-resampled (dark shading) and parameter-resampled IPM models (light shading). Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class; the marginal plots indicate the distribution of these relative proportions for each class of IPM in each habitat. Note that both the scatterplots and marginal plots exclude transient dynamics (iterations 1-30)."

@@

```{r aic, results='asis'}
#| label: tbl-aic
#| tbl-cap: "Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. <i>'edf'</i> is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
#| tbl-pos: H
#| 
aic_tbl_df %>%
  group_by(habitat, model) %>% 
  # mutate(rank = row_number()) %>% 
  ungroup() %>% 
  mutate(habitat = toupper(habitat),
         env_effect_incl = ifelse(
           env_effect_incl != "DLNM",
           to_sentence_case(as.character(env_effect_incl)),
           as.character(env_effect_incl)
         )
  ) %>% 
  mutate(
    model = fct_relevel(
      model,
      c("surv", "size", "flwr", "surv_sdlg", "size_sdlg")
    ),
    model = fct_recode(
      model,
      "Survival"          = "surv",
      "Growth"            = "size",
      "Flowering"         = "flwr",
      "Seedling survival" = "surv_sdlg",
      "Seedling growth"   = "size_sdlg"
    )
  ) %>%

@@

  arrange(habitat, model) %>% 
  mutate(habitat=c(rep("Continuous Forest",15),rep("Forest Fragments",15))) %>% 
  mutate(model=(rep(c("Survival","","","Growth","","","Flowering","","","Seedling survival","","","Seedling growth","",""),2)))  %>%
  select(
    "Habitat" = "habitat",
    "Vital Rate" = "model",
    "Effect of Environment" = "env_effect_incl",
    #"Model" = "RHS",
    "\\textit{edf}" = "df",
    "\u0394AIC" = "dAIC"
  )  %>% 
  mutate(Habitat = str_replace_all(
    Habitat,
    c(
      "Continuous Forest" = "",
      "Forest Fragments" = ""
      )
  )) %>% 
  kable(digits = 2,
  format = "latex",
  align = "cllcc",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  linesep = ""
) %>%
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 10,
    position = "center"
  ) %>% 
# %>% 
  # collapse_rows(columns = 1:2, valign = "top")
  column_spec(1, width = "10em") %>% 
  pack_rows("Continuous Forest", 1, 15) %>%
  pack_rows("Forest Fragments", 16, 30)

@@

lambda_table %>% 
  mutate(habitat = toupper(habitat)) %>% 
  mutate(across(c(est, lower, upper), ~format(round(.,4)))) %>% 
  # mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
  mutate(CI = glue::glue("({lower}, {upper})")) %>%
  mutate(ipm = str_replace_all(
    ipm,
    c(
      "det" = "Deterministic",
      "stoch" = "Stochastic, kernel-resampled",
      "dlnm" = "Stochastic, parameter-resampled"
    )
  )) %>% 
  # separate(lambda,c("lambda","CI"),sep=" [(]") %>% 
  select(IPM = ipm, 
         Habitat = habitat, 
         # "$\\lambda$" = lambda) %>% 
         "$\\lambda$" = est,
         "95\\% CI (Lower, Upper)"=CI) %>%
  mutate(IPM = str_replace_all(
    IPM,
    c(
      "Deterministic" = "",
      "Stochastic, kernel-resampled" = "",
      "Stochastic, parameter-resampled" = ""
      )
  )) %>%  
  kable(digits = 2,
  format = "latex",
  align = "lccc",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  linesep = ""
) %>% 
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  ) %>% 
  # collapse_rows(columns = 1:2, valign = "top")
  column_spec(1, width = "15em") %>% 
  pack_rows("Deterministic", 1, 2) %>%
  pack_rows("Stochastic, kernel resampled", 3, 4) %>% 
  pack_rows("Stochastic, parameter-resampled", 5, 5)

@@

stats_table %>%
  # select(-`Vital Rate`) %>%
  rename(`Stage`=`Vital Rate`,
         "\\textit{df}"=df,
         "\\textit{P}"=`p value`) %>%  
  mutate(`Stage`= " ") %>% 
  # mutate(Comparison="KR v PR") %>% 
  # relocate(Comparison,.before=1) %>% 
kable(digits = 5,
  format = "latex",
  align = "lcrrcr",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  # caption = "Test comparing variances of vital rates in stochastic ipms.",
  linesep = ""
) %>%
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  ) %>%
  # collapse_rows(columns = 1:2, valign = "top")
  column_spec(1, width = "8em") %>%
  # collapse_rows(columns = 1:2,valign="top")
  pack_rows("Seedling", 1, 2) %>%
  pack_rows("Pre-reproductive 1", 3, 4) %>%
  pack_rows("Pre-reproductive 2", 5, 6) %>%
  pack_rows("Reproductive 1", 7, 8) %>%
  pack_rows("Reproductive 2", 9, 10)
